\documentclass[12pt, a4paper]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage[section]{placeins}
\usepackage{paralist}
\usepackage{fullpage}
\usepackage{glossaries}
\usepackage[hidelinks]{hyperref}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}

\newacronym{GSM}{GSM}{Global System for Mobile Communications}
\newacronym{NSA}{NSA}{National Security Agency}
\newacronym{SELinux}{SELinux}{Security-Enhanced Linux}
\newacronym{SIM}{SIM}{Subscriber Identity Module}
\newacronym{ICO}{ICO}{Information Commissioner's Office}

\newcommand*{\titleGM}{\begingroup
\hbox{ 
\rule{1pt}{\textheight} 
\hspace*{0.02\textwidth} 
\parbox[b]{0.97\textwidth}{ 

{\noindent\Huge\bfseries Mobile Devices have made us the most monitored individuals in history}\\[2\baselineskip] % Title
{\large \textit{SEM2220 - Assignment 4}}\\[4\baselineskip] % Tagline or further description
{\Large \textsc{080932464}} % Author name

\vspace{0.5\textheight} 
}}
\endgroup}


\begin{document}
\titleGM 
\tableofcontents
\newpage

\section{Introduction}

There has been a lot of news in the press in the last year about surveillance
programs run by many countries security agencies, particularly focused on the
collection of phone records by the \gls{NSA}\cite{greenwald2013nsa}.

It comes as no surprise that this has brought about the question: 

\begin{quote}
``Have mobile devices made us the most monitored individuals in history?''
\end{quote}

To fully understand the depth of this question, one must consider the technical
methods which are employed to monitor mobile devices and how these methods
differ across different mobile platforms. In hand with this, one must
investigate the methods for preventing unwanted monitoring on mobile devices
and how effective these methods really are.

Finally there are many legal and professional issues which govern the
collection of data from mobile devices; but there are also a host of ethical
and social issues which legal and professional element cannot, necessarily,
cover.

Of course monitoring of data, especially from mobile devices, doesn't have to
be used for nefarious purposes. Applications commonly use data mining 
techniques to provide a better experience for their users and cloud services 
are so popular that imagining a world where all a users data is stored solely 
on a single device is almost impossible.

\newpage
\section{Technical Methods for Monitoring Mobile Devices}

Because of the structure of the \gls{GSM}, any mobile device which accesses
the mobile network can be tracked to, at worst, the nearest cell tower. In 
fact the \gls{NSA} is known to use this data to track targets and identify 
possible accomplices\cite{classified2012co-travel}. There are also other
stories of UK public organisations performing similar
actions\cite{gallager2011met}\cite{hopkins2013gchq}.

Though this is a worrying concept, there is some important information that the
U.S. has given in \cite{clapper2013facts} which denies that any information
collected cannot be used to target any individual without a prior specific and
documented reason to do so. In the case of monitoring of non-U.S. citizens
there must also be a agreement with the country of their citizenship.

It is the monitoring which the users agrees to have performed upon them that is
perhaps more worrying.

Mobile devices, especially tablets and smart phones, have a range of sensors
which third-party applications may be able to access. These sensors typically
include a camera, microphone, WiFi antenna and GPS system. As technology
progresses the accuracy of these sensors is increasing and other sensors such
as accelerometers are becoming more prevalent.

It could, therefore, be easy for an attacker to write a malicious application
which simply monitored the GPS location of a person or recorded their
conversations through the microphone. Or even use any of these systems combined
to gather a lot of sensitive data.

There have been cases which have been formally
investigated\cite{ftc2013flashlight} in which legitimate applications collect
data from users on the side, which are then sold to other companies for
advertisement revenue.

However, a lot of applications have legitimate uses for these sensors. Most
applications require network access to load data from external servers. More
specific applications, which perform tasks like route tracking, will require
the use of more than one of these sensors.

\newpage
\section{Protection Against Unwanted Monitoring}

Unfortunately, due to the structure of the \gls{GSM} network, as long as a
mobile device is on and allowed to connect to the network it can be tracked.
With the addition of a \gls{SIM} card it can even be tied to a specific
individual. Under \cite{directive2006/24/EC}

Modern mobile platforms typically have some sort of security system and ways in
which applications can be given access to secure elements of the mobile device.
Most of these platforms sandbox third-party application allowing the kernel to
control the access to the various facilities of the device.

\subsection{Android Security Facilities}

\cite{google2013security} describes the methods that the Android system uses to
secure the environment. The main advantage Android gains in this area is that
uses \gls{SELinux}, which confines the access of programs based on different
policies.

Android adds many features on top of this, the most applicable of these to the
question of monitoring is the permission model. For a third party application
to access protected APIs such at GPS positioning or network connections they
must implicitly define which of these elements they require in the manifest
file of the application.

This information is displayed to the user when they first install the
application and when the application updates to add new access to secure APIs.

There are also certain APIs which are not available to third-party 
applications, but which may be used by pre-installed applications if they are
signed as part of the OS.

Because applications are sandboxed from one another it should be an impossible
task to gain confidential information from another application, unless it
provides it somehow.

\subsubsection{Additional Security Features Provided by the Google Play Store}

Obviously, telling legitimate use of secure API elements from illegitimate use
is not an easy task, especially for automated systems.

In the mobile ecosystem, the typical method of performing this is to have a
trusted place, where developers can distribute their applications. In the case
of Android this is usually the Google Play store, however other stores such as
the Amazon Appstore do also exist.

The maintainers of such places can then regulate the applications accepted,
monitor them for any malicious activities and remove them where needed. 
Google even added a service in 2012 which scanned the Play store for
applications with a malicious intent by running them on a range of emulated
services\cite{lockheimer2012android}.

\nocite{lockheimer2012android}

\subsection{iOS Security Facilities}
\cite{apple2012security} defines the methods that the iOS system uses to secure
the environment. Similar to Android iOS third-party applications are sandboxed 
from each other, although a lot of this is done by the iOS runtime. All
applications have their own file spaces which can only be accessed through the
iOS defined API.

Access to secure API elements is defined through entitlements; signed key-value
pairs specific to applications. However, it should be noted that these
entitlements are not required to access elements such as the GPS or microphone
and are more typically used by system applications and daemons to perform tasks
that would typically require root access.

The iOS ecosystem is heavily reliant on the Apple App
Store to provide security against the malicious
collection of data from iOS applications.

%TODO how do they do this?

%\subsection{Windows 8 Phone Security Features}



\newpage
\section{Implications of Mobile-based Monitoring}

This section will focus on the four main implications of developing software
which gathers data from mobile devices; legal, professional, social and
ethical.


\subsection{Legal Implications}

There are several laws which govern the legalities of computer software and the
storage of personal data. The notable laws which apply to the monitoring of
data gathered by mobile devices are:

\begin{itemize}
\item Data Protection Act\cite{dpa1998}
\item Computer Misuse Act\cite{cma1990}
\end{itemize}

Other laws may still apply to the process of developing the software which
performs this, but are less relevant to the actual collection of data.


\subsubsection{Impacts of the Data Protection Act}
Monitoring data undoubtedly also means storing data and any computerised data
falls under the Data Protection Act; therefore, any monitoring is going to be
affected by the Data Protection Act.

\gls{ICO} has released guidelines\cite{ico2012cloud} on the use of cloud 
computing and how it can be kept legal within the Data Protection Act. These
mainly state that it is essential to have a contract in place with a trusted
cloud storage provider that will adhere to UK law. This might be especially
difficult with many cloud providers being non-UK or even -EU based.

The EU has made agreements with the US government for a safe harbour for
private data that could be stored on US servers\cite{eu2000safe}, this does
help with some issues that storing data on a non-EU cloud, but given the way in
which clouds tend to operate, there is no non-contractual way in which this can
be guaranteed.

\subsubsection{Impacts of the Computer Misuse Act}
The Computer Misuse Act does not affect the use of mobile devices for monitoring
purposes as much as one would suspect. The main issue comes that section 1 of
this act defines misuse as ``the unauthorised access to computer
material''\cite{cma1990}. Sections 2 and 3 also define other forms of misuse, 
but as monitoring would fall under the access of computer material, they do not
apply.

The problem comes that the access could be implied simply by the user installing
the application. Most applications will also have terms and conditions or will
be distributed in such a way that the user is accepting the access to
information providing by sensors before they install the application (e.g. the
Android permission system).

Obviously, if an application could hide the fact it was collecting data from a
permission system and does not disclose this use in terms and conditions, then
an argument could be made that it is breaking this act.


\subsection{Professional Implications}

Section 1.a of the BSC Code of Conduct states that ``[You shall] have due regard
for the public health, privacy, security and wellbeing of others and the
environment''\cite{bsc2011conduct}. This would imply that you must secure any
data collected through the monitoring of mobile devices. Having due regard for
an individuals privacy, might also imply that the results of the monitoring must 
be kept private, and not, for example, sold to another company to generate
revenue.

Section 2.f also discusses avoidance of malicious and negligent action which
could imply that any professional should not be developing a system which
monitors mobile devices for malicious purposes.


\subsection{Social Implications}

There is a lot of good that can come from the monitoring of mobile devices and
as mentioned there may be many good reasons for application developers to
perform monitoring. Android 4.0 added the ability to monitor roaming data 
usage, so a user could set a warning message when they reach a certain amount 
of usage, and to cap their usage at a different limit\cite{android2011ics}; 
this feature gives the user more control over the use of their mobile phone, 
and prevents them from spending more than they need to on their phone bill.

Given that less that 1\% of all UK premises receive no 3G connection from any
provider and that only around 23\% of all geographical areas in the UK receive
no 3G connection from any provider\cite{offcom2013infastructure} and that the
figures for 2G connections are even less, it is difficult to not be connected
to a form of network. With the recent adoption of 4G networks these figures are
likely to decrease even further.

Given that a lot of services provide synchronisation to the cloud, it is
difficult to escape the pressures of work when emails can be delivered to a
mobile device so easily. There have been studies into the affects of using
email constantly against taking vacations from emails leading to reduced stress
and increased productivity\cite{mark2012pace}. Similar studies have also been
performed on the mental welfare through the increased use of
Facebook\cite{kross2013facebook}, where increased use lead to the increased
chances of depression in young adults.

Monitoring data could be one of the possible reasons why there are such
stresses in modern day life, but it could also allow stress to be reduced by
limiting access to certain applications or suggesting improvements to fitness
and diet.


\subsection{Ethical Implications}
The final consideration is that of ethics; it may be legal, professional and
socially acceptable to develop an application which monitors some information
from a users device, but the purpose or reasons for doing so may not be
ethical.

Of course ethics are a very personal viewpoint, to help standardise ethics, the
Computer Ethics Institute has released ten commandments\cite{ceiXcommandments} 
which, if followed, would deem a application ethical.

The author has his own views on what makes an application ethical; using Android
as an example, there should be a reasonable explanation as to why an
application wants to access a protected API and this reason should be presented
to the user, rather than just the protect API elements an application has.

\newpage
\section{Conclusions}

Yes, mobile devices have probably aided in being the most monitored individuals
in history. To say mobile devices are the only reason for this would be an
unsound argument; though location could be seen as more sensitive than internet
history, for example.

Mobile devices just add a lot more ways in which individuals can be monitored;
however, the access to these systems is more strictly controlled compared to
the standard PCs or laptops. Whilst this is a good step in the right direction,
there are also additional items which cannot be so strictly controlled as it is
stored one systems external to the device (i.e. the \gls{GSM} network).

\newpage
\bibliographystyle{IEEEtran}
\bibliography{citations}

\end{document}
